---
title: "Modeling Transcription-Accessibility Coupling in Spermatogenesis"
subtitle: "Figure 5A Analysis: RS to ES Transition"
output: html_document
---

```{r setup, include=FALSE}
# Load essential libraries for modeling, statistics, and visualization
library(ggplot2)   # Visualization
library(lme4)      # Mixed-effects modeling
library(lmerTest)  # Satterthwaite's d.f. for p-values
library(emmeans)   # Post-hoc trend analysis
library(dplyr)     # Data manipulation
library(car)       # Hypothesis testing
library(ggtext)    # Rich text for ggplot labels
```

# 1. Data Ingestion and Factors 
The primary dataset consists of ATAC-seq signal and transcriptomic data (txn) across four developmental stages.
``` {r}
# Read regression data and standardize factor levels to developmental order
df <- read.csv('txn_atac_AT_TSS_regression_data_ref_GRCm38.81.csv')
df$stage <- factor(df$stage, levels = c("early_RS", "int_RS", "late_RS", "early_ES"))

# Standardize column naming
names(df)[names(df) == "mean_ATAC"] <- "ATAC"
```

# 2. QC & Filtering
To ensure the model captures biological coupling rather than extreme outliers or background noise, specific gene filters are applied.
## Remove early_ES highly transcribed genes
Protamines and transition proteins (Prm1, Prm2, Tnp1, Tnp2) are excluded as they undergo massive transcriptional surges in early Elongating Spermatids (ES).
``` {r}
genes_to_remove <- c("Prm1", "Prm2", "Tnp1", "Tnp2")
df_filtered <- df[!((df$gene_name %in% genes_to_remove)), ]
```


## Activity based filtering
Genes in the bottom 25% of mean transcription are removed to focus on transcriptionally active loci.
``` {r}
# Identify genes above the bottom quartile of mean transcription
gene_stats <- df %>%
  group_by(gene_name) %>%
  summarize(mean_txn = mean(txn))

cutoff_val <- quantile(gene_stats$mean_txn, 0.25)

df_clean <- df %>%
  group_by(gene_name) %>%
  filter(mean(txn) > cutoff_val) %>%
  ungroup()

# Final dataset excluding both low-activity and high-leverage genes
df_final <- df_clean[!((df_clean$gene_name %in% genes_to_remove)), ]
```


# 3. Mixed Effect Modeling
We employ a Linear Mixed-Effects Model (LMM) to model transcription and accessibility
coupling. 

``` {r}
# Fit model with interaction between transcription and stage
# A random intercept is assigned to each gene_name
m_log <- lmer(ATAC ~ log1p(txn) * stage + (1|gene_name), 
              data = df_final, 
              na.action = na.exclude)

summary(m_log)
```


# 4. Model Diagnostics

``` {r}
par(mfrow = c(1, 2))
# Residuals vs Fitted: Checks for non-linear patterns or heteroscedasticity
plot(fitted(m_log), resid(m_log),
     main = "Residuals vs Fitted",
     xlab = "Fitted Values", ylab = "Residuals",
     pch = 20, col = rgb(0,0,0,0.1))
abline(h = 0, col = "red")

# Q-Q Plot: Checks for normality of residuals
qqnorm(resid(m_log), pch = 20)
qqline(resid(m_log), col = "red")
```
Visual inspection indicated minor heteroscedasticity; however, parameter estimates in LMMs remain unbiased even under violation of homoscedasticity, though precision may be affected (Schielzeth 2020)

# 5. Post-hoc Slope Analysis
We extract the marginal slopes (trends) for each stage to quantify the strength of coupling.
``` {r}
# Calculate slope estimates for log(txn) per stage
emm_slopes <- emtrends(m_log, specs = "stage", var = "txn")
print(emm_slopes)

# Pairwise comparisons of slopes between all stages
pairs(emm_slopes)
```

# 6. Joint Hypothesis Test
To determine if the transcription-accessibility coupling in the early ES stage is distinct from the RS stages, we perform a Wald Test (Joint Hypothesis Test). This test evaluates whether the interaction terms involving the early_ES stage are significantly different from the reference stage and other RS stages simultaneously.
```{r}
# Perform a Joint Wald Test to compare the early_ES slope against all RS stages
# This checks if the 'uncoupling' observed in ES is statistically significant across the transition
joint_test <- linearHypothesis(m_log, c(
  "log1p(txn):stageearly_ES = 0",                              # ES vs. Reference (early_RS)
  "log1p(txn):stageint_RS - log1p(txn):stageearly_ES = 0",     # ES vs. int_RS
  "log1p(txn):stagelate_RS - log1p(txn):stageearly_ES = 0"     # ES vs. late_RS
))

print(joint_test)
```

# 7. Visualizations

## Fitted coupling trends
This plot visualizes the regression lines for each stage, highlighting the uncoupling in the Early ES stage.
```{r}
# Generate prediction data
plot_data_trends <- emmip(m_log, stage ~ txn, cov.reduce = range, CIs = TRUE, plotit = FALSE)
plot_data_trends$stage_label <- factor(plot_data_trends$stage,
                                       levels = c("early_RS", "int_RS", "late_RS", "early_ES"),
                                       labels = c("Early RS", "Int RS", "Late RS", "Early ES"))

p <- ggplot(plot_data_trends, aes(x = log1p(txn), y = yvar, color = stage_label, fill = stage_label)) +
  geom_ribbon(aes(ymin = LCL, ymax = UCL), alpha = 0.1, color = NA) +
  geom_line(linewidth = 1) + 
  #scale_color_manual(values = c("Early RS" = "grey80", "Int RS" = "grey40", 
  #                              "Late RS" = "grey10", "Early ES" = "#D55E00")) +
  #scale_fill_manual(values = c("Early RS" = "grey80", "Int RS" = "grey40", 
  #                              "Late RS" = "grey10", "Early ES" = "#D55E00")) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(y = "Fitted ATAC Signal\n(GC-residual)", x = "Transcription Level (log)", color = "Stage", fill = "Stage") +
  theme_classic(base_size = 14)

ggsave("/Users/zapell/Library/CloudStorage/Dropbox-UniversityofMichigan/Zachary Apell/HiC_collaborations/figure_plots/final_figs/fig5a_InteractionPlot_Dark2_color.png", plot = p, width = 4, height = 3.5, units = "in", dpi=300)
```


## Slope Estimates with confidence intervals
A direct comparison of the coupling strength (slope) across development.
```{r}
plot_slopes <- as.data.frame(emm_slopes)

#ggplot(plot_slopes, aes(x = stage, y = txn.trend, color = stage == "early_ES")) +
#  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
#  geom_pointrange(aes(ymin = asymp.LCL, ymax = asymp.UCL), size = 1, linewidth = 1) +
#  scale_color_manual(values = c("FALSE" = "grey40", "TRUE" = "#D55E00")) +
p2 <- ggplot(plot_slopes, aes(x = stage, y = txn.trend, color = stage)) + 
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_pointrange(aes(ymin = asymp.LCL, ymax = asymp.UCL), size = 1, linewidth = 1) +
  scale_color_brewer(palette = "Dark2", labels = c("Early RS", "Int RS", "Late RS", "Early ES")) +
  scale_x_discrete(labels = c("early_RS" = "Early RS", "int_RS" = "Int RS", 
                              "late_RS" = "Late RS", "early_ES" = "Early ES")) +
  labs(y = "Slope Estimate (\u03B2) \u00B1 95% CI", x = NULL) +
  theme_classic(base_size = 14) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("/Users/zapell/Library/CloudStorage/Dropbox-UniversityofMichigan/Zachary Apell/HiC_collaborations/figure_plots/final_figs/fig5a_slope_estimation_error_bar_filtered2_log_Dark2.png", plot = p2, width = 4, height = 3.5, units = "in", dpi=300)
```



## Pairwise Comparisons and Contrast Visualization
While the joint test confirms an overall difference, pairwise comparisons identify specific shifts between developmental stages. We visualize these differences (Δβ) using a forest-style contrast plot.
```{r}
# 1. Generate all pairwise contrasts for the slope estimates
contrasts_all <- as.data.frame(pairs(emm_slopes, infer = c(TRUE, TRUE)))

# 2. Format p-values (using 'p < 0.001' notation for small values)
contrasts_all$p_label <- ifelse(contrasts_all$p.value < 0.001, "p < 0.001",
                         paste0("p = ", sprintf("%.3f", contrasts_all$p.value)))

# 3. Define a consistent color scheme: Greyscale for RS, Orange for ES
#color_map <- c(
#  "early_RS"    = "grey80", "int_RS"      = "grey40",
#  "late_RS"     = "grey10", "early_ES"    = "#D55E00"
#)
# Load the package at the top of your script or chunk
library(RColorBrewer)

# Automatically pull the first 4 colors from the Dark2 palette
pal <- brewer.pal(n = 4, name = "Dark2")

# 3. Define a consistent color scheme automatically
color_map <- c(
  "early_RS" = pal[1], "int_RS"   = pal[2],
  "late_RS"  = pal[3], "early_ES" = pal[4]
)

# Helper function to clean stage names for labels
clean_name <- function(x) {
  x <- gsub("_", " ", x)
  tools::toTitleCase(x)
}

# 4. Generate HTML-styled labels for the X-axis using ggtext
contrasts_all$styled_label <- sapply(seq_len(nrow(contrasts_all)), function(i) {
  x <- as.character(contrasts_all$contrast[i])
  parts <- unlist(strsplit(x, " - "))
  
  col1 <- ifelse(parts[1] %in% names(color_map), color_map[parts[1]], "black")
  col2 <- ifelse(parts[2] %in% names(color_map), color_map[parts[2]], "black")
  
  paste0("<span style='color:", col1, "'>", clean_name(parts[1]), "</span><br>",
         "<span style='color:", col2, "'>", clean_name(parts[2]), "</span>")
})

# Sort contrasts by the magnitude of the estimate for visual clarity
contrasts_all <- contrasts_all %>% arrange(estimate)
contrasts_all$styled_label <- factor(contrasts_all$styled_label, 
                                     levels = unique(contrasts_all$styled_label))

# plot
ggplot(contrasts_all, aes(x = styled_label, y = estimate)) +
  # Zero line represents no difference in coupling strength
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  
  # Error bars and point estimates
  geom_pointrange(aes(ymin = asymp.LCL, ymax = asymp.UCL), 
                  color = "black", size = 0.8, linewidth = 0.8) +
  
  # Annotated p-values
  geom_text(aes(y = asymp.UCL, label = p_label), 
            vjust = -0.5, color = "black", size = 3.5) +
  
  labs(
    title = "All Pairwise Slope Comparisons",
    x = "Contrast (Top Stage - Bottom Stage)",
    y = "Difference in Slope Estimate (\u0394\u03B2)"
  ) +
  
  # Expand Y-axis to prevent p-value labels from being cut off
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.15))) +
  
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_markdown(lineheight = 1.2), # Enables HTML rendering
    axis.text.y = element_text(color = "black"),
    axis.title.x = element_text(size = 12, margin = margin(t = 10)) 
  )
```
